---
title: Boxplot with quartiles
author: FS
date: '2018-09-23'
slug: plotting-quartiles
categories: [R]
tags: []
header:
  caption: ''
  image: ''
---

Boxplots are one of the best ways to represent distributions. Below, there is some
code to create a costum boxplot by determining the pecentiles of the distribution that the wiskers and the box represent. In the example, I plotted the quartiles of the distribution of weight as a function of group from the dataset *PlantGrowth*.

First, we create a function to extract quartiles. The values in **probs** represent the percentiles of the distributions, which then will be plotted. Then, we examine the quartiles of the distribution of weight for each group (i.e., ctrl, trt1, trt2).

```{r function quartiles, echo=TRUE, results='markup'}
# Create the function quartiles
quartiles <- function(x) {
  r <- quantile(x, probs=c(0, 0.25, 0.5, 0.75, 1))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}

# Quartiles of weight per group
by(PlantGrowth$weight, INDICES = PlantGrowth$group, FUN = quartiles)

```

Finally, the boxplot using ggplot. I also added the diamond shape to represent the mean and the individual values jittered around the boxplot. 

```{r plot, results='markup', echo=TRUE, warning=FALSE}

# Libraries
library(ggplot2)

# Plot
ggplot(data = PlantGrowth, aes(y=weight, x=group, width=0.3)) +
  stat_summary(fun.data = quartiles, geom="boxplot", size=0.7, fill="white") +
  stat_summary(fun.y = mean, geom = "point", shape=18, size=5) +
  geom_jitter(width = 0.1, height = 0, alpha=0.3, size=3) +
  theme_classic(base_size = 18) +
  guides(fill=guide_legend(title="Triplet")) +
  theme(axis.title.x = element_text(face="bold"),
        axis.title.y = element_text(face="bold"),
        legend.title = element_text(face="bold"),
        axis.text.x = element_text(colour = "Black"),
        axis.text.y = element_text(colour = "Black"),
        axis.ticks.x = element_line(colour = "Black", size = 0.2),
        axis.ticks.y = element_line(colour = "Black", size = 0.2),
        axis.line.x = element_line(size = 0.2),
        axis.line.y = element_line(size = 0.2),
        panel.border = element_rect(size = 0.2,color="black", fill=NA),
        plot.title = element_text(face="bold", hjust = 0.5),
        panel.grid.major = element_line(size = (0.1), colour="grey"),
        panel.grid.minor = element_blank()) +
  xlab("Group") +
  ylab("Weight") +
  ggtitle("Boxplot quartiles")

```